---
title: "SMPA - Grouping Assignment"
author: "Jenna Baer"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)

#load data
payments <- readRDS("data/trump_campaign_payments.rds")

```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.   

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  I would look to see the total disbursements of each month to determine a pattern in spending. I want to know if the campaign is increasing the amount they spend over time,as election day approaches.

2.  I would also want to determine who is receiving the most money from the campaign (which company/ person)?

3.  Finally, I would like to know what cities the campaign is infusing cash into with their business.

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

In 2017,$17082338 was spent. In 2018, $37212716	was spent. I found these numbers by grouping the data based upon year (separating it out). And then was able to count the totals by summarizing based on the total sums of disbursements for each year.

```{r}

payments %>%
  group_by(disbursement_year) %>%
  summarise(total_disbursement = sum(disbursement_amount))


  





```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

October 2018 had the highest spending of $9738567.6. I found this by again grouping by disbursement year but also incorporating disbursement month to differentiate between the various months. I again used the summarize function to count the sums of disbursment amounts.
```{r}

payments %>%
  group_by(disbursement_month, disbursement_year) %>%
  summarise(total_disbursement = sum(disbursement_amount))




```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?
 
AMERICAN MADE MEDIA CONSULTANTS, LLC was the highest paid at $9772583.64. I figured this out by grouping by recipient name this time (to combine all of the charges into each respective recipient). From there I used the summary function to count the sums of the disbursement amounts.
```{r}


payments %>%
  group_by(recipient_name) %>%
  summarise(total_disbursement = sum(disbursement_amount)) %>%
  arrange(desc(total_disbursement))




```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

They were paid the most for placed media, $4334158.34. I figured this out by filtering for the recipient name variable and inserting AMERICAN MADE MEDIA CONSULTANTS, LLC. I then grouped based on disbursement description and summarized based on the sum total of disbursement amount for each description.
```{r}

payments %>%
  filter (recipient_name == "AMERICAN MADE MEDIA CONSULTANTS, LLC") %>%
  group_by(disbursement_description) %>%
  summarise(total_disbursement = sum(disbursement_amount))




```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

Merchant fees was the most common category (1002 times). I determined this by grouping based on description and then summarizing to get a straightfoward count. I then arranged my data in descending order. 
```{r}

payments %>%
  group_by(disbursement_description) %>%
  summarise( count = n ()) %>%
  arrange(desc(count))




```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

JONES DAY	$4390351.66	
LAROCCA HORNIK ROSEN GREENBERG & BLAHA LLP	$404774.48	
KASOWITZ, BENSON, TORRESS & FRIEDMAN	$322614.29	
THE LAW OFFICES OF ALAN S. FUTERFAS	$287924.00	
MCDERMOTT WILL & EMERY	$276278.06

I figured this out by filtering for legal consulting, then grouping based on the recipient name (because a lot of recipients received multiple payments). From there I calculated the totals that each legal consulting firm received by using the summarise function. I then arranged these totals in descending order to get the top 5.
```{r}

payments %>%
  filter(disbursement_description == "LEGAL CONSULTING") %>%
  group_by(recipient_name) %>%
  summarise(total_disbursement = sum(disbursement_amount)) %>%
  arrange(desc(total_disbursement))







```


Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

Top 5 individuals:
Michael GLassner: $511280.00	
SEAN	DOLLMAN:	$389541.09	
DARREN	CENTINELLO:	$214934.25	
CHRISTOPHER	CARR:	$160000.00	
JOHN	PENCE:	$104244.03

I filtered for only individuals so I could parse out the top 5 highest paid. From there I grouped by first and last name to condense the people received multiple payments. I then counted up the payments by using the sum function to add up the disbursement amounts. From there I arranged in descending order to figure out the top 5.
```{r}

payments %>%
  filter(entity_type_desc == "INDIVIDUAL") %>%
  group_by(payee_first_name, payee_last_name) %>%
  summarise(total_disbursement = sum(disbursement_amount)) %>%
  arrange(desc(total_disbursement))






```


For the indvidual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

Michael Glassner was paid $499000.00 in the payroll category. I calculated this by filtering for payee last name and substituting in Glassner. I then grouped by disbursement description to combine all of the similar categories. From there I added up the total disbursements by using the summarise function.



```{r}


payments %>%
  filter(payee_last_name == "GLASSNER") %>%
  group_by(disbursement_description) %>%
  summarise(total_disbursement = sum(disbursement_amount))




```
